---
title: "Data wrangling"
author: "Fonti Kar"
date: "2024-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load dependencies

```{r}
# install.packages("pacman")
pacman::p_load(tidyverse, janitor, snakecase, arrow)
```

### Read in data

```{r}
raw <- read_csv("data/15042024-095250-report.csv", name_repair = "minimal")
```

### Deal with two header problem

```{r}
# Column names
higher_level_headers <- colnames(raw) |> snakecase::to_snake_case()

# First row
informed_headers <- c(raw[1,]) |> unname() |> unlist() |> snakecase::to_snake_case()

# listed_name is non unique, need to paste on act/body name to listed_name
informed_headers |> str_subset("^listed_name$")
informed_headers |> str_subset("act")

# Hard code solution for now using indices
acts_cols <- informed_headers[31:45]

# Drop the listed_name
acts_names <- acts_cols |> str_subset("^listed_name", negate = TRUE)

# Paste act names to listed_name
fixed_listed_name <- paste(acts_names, "listed_name", sep = "")

# Replace plain old listed_name with new improved ones
informed_headers[informed_headers |> str_which("^listed_name$")] <- fixed_listed_name

informed_headers

# Paste these together
combined_headers <- paste(higher_level_headers, informed_headers, sep = "-")

# Remove 1st row of names
cleaned_headers <- raw[-1,]

# Rename column headers
colnames(cleaned_headers) <- combined_headers

colnames(cleaned_headers)
```

### Arrange the taxonomic data together, remove `epc_act`, remove `-NA`

```{r}
threatened_species <- cleaned_headers |> 
  select(-starts_with("epc_act")) |> 
  select(starts_with("current_name"), starts_with("taxonomic_data"), starts_with("state_territory"))

colnames(threatened_species)
```

### Ditch long prefixes in names

```{r}
shortened_names <- colnames(threatened_species) |> 
  str_remove("current_name_and_sprat_id-") |> 
  str_remove("state_territory_and_international_threatened_species-") |> 
  str_remove("taxonomic_data-")


colnames(threatened_species) <- shortened_names
colnames(threatened_species)
```

### Condense the acts/bodies together

# Status 

```{r}
status_by_body <- threatened_species |> 
  select(taxon_id:taxon_group, act_nc_act, wa_wc_act, nsw_tsc_act_and_fm_act, 
                        qld_nc_act, tas_tsp_act, sa_npw_act, vic_ffg_act_advisory_lists, 
                        iucn_red_list, nt_tpwc_act) |> 
  pivot_longer(cols = c(act_nc_act:nt_tpwc_act), 
               names_to = c("threatened_species_body"),
               values_to = "conservation_status") |> 
  relocate(threatened_species_body, conservation_status, .after = taxon_group)
```

# Listed names

```{r}
listed_names_by_body <- threatened_species |> 
  select(taxon_id:taxon_group, ends_with("listed_name"), ends_with("listed_names")) |> 
  pivot_longer(cols = c(act_nc_act_listed_name:iucn_red_list_listed_names), 
               names_to = c("threatened_species_body"),
               values_to = "listed_names") |> 
  relocate(threatened_species_body, listed_names, .after = taxon_group)

# Fix the values in threatened_species_body
listed_names_by_body_cleaned_body <- listed_names_by_body |> 
  mutate(threatened_species_body = str_remove_all(listed_names_by_body$threatened_species_body, "_listed_name")) 
```

### Condense the listed_names together

```{r}
nrow(status_by_body)
nrow(listed_names_by_body)

threatened_species_by_body <- left_join(status_by_body, listed_names_by_body_cleaned_body) 
```

### Invertebrates only Phylums

```{r}
threatened_species_by_body |> 
  tabyl(phylum)

threatened_inverts <- threatened_species_by_body |> 
  filter(kingdom == "Animalia") |> 
  filter(!phylum == "Chordata") 
```

### Exclude least concern and data deficient in IUCN

```{r}
threatened_inverts$conservation_status |> 
  tabyl()

threatened_inverts |> 
  filter(! conservation_status == "Data Deficient" & ! conservation_status == "Least Concern") |> # First filter
  filter(! str_detect(conservation_status, regex("deficient", ignore_case = TRUE))) |> # Instances where deficient is mentioned
  filter(! str_detect(conservation_status, regex("least", ignore_case = TRUE))) |> # Instances where least is mentioned
  filter(! is.na(conservation_status)) |> # Missing conservation status
  tabyl(conservation_status)

ditched_deficient_least_concern <- threatened_inverts |> 
  filter(! conservation_status == "Data Deficient" & ! conservation_status == "Least Concern") |> # First filter
  filter(! str_detect(conservation_status, regex("deficient", ignore_case = TRUE))) |> # Instances where deficient is mentioned
  filter(! str_detect(conservation_status, regex("least", ignore_case = TRUE))) |> # Instances where least is mentioned
  filter(! is.na(conservation_status))
```

### Fix up values in conservation status

```{r}
ditched_deficient_least_concern |> 
  tabyl(conservation_status)

ditched_deficient_least_concern |> 
  mutate(conservation_status_cleaned = 
           if_else(conservation_status == "Critically endangered", "Critically Endangered", conservation_status)
         ) |> 
  mutate(conservation_status_cleaned = 
           if_else(conservation_status_cleaned == "Near Threatened, Near Threatened","Near Threatened", conservation_status_cleaned)
         ) |>
   mutate(conservation_status_cleaned = 
           if_else(conservation_status_cleaned == "Presumed extinct","Presumed Extinct", conservation_status_cleaned)
         ) |> 
           tabyl(conservation_status_cleaned)


cleaned_status <- ditched_deficient_least_concern |> 
  mutate(conservation_status_cleaned = 
           if_else(conservation_status == "Critically endangered", "Critically Endangered", conservation_status)
         ) |> 
  mutate(conservation_status_cleaned = 
           if_else(conservation_status_cleaned == "Near Threatened, Near Threatened","Near Threatened", conservation_status_cleaned)
         ) |>
   mutate(conservation_status_cleaned = 
           if_else(conservation_status_cleaned == "Presumed extinct","Presumed Extinct", conservation_status_cleaned)
         )  
```

### Export

```{r}
write_csv(cleaned_status, file = paste0("output/", paste(Sys.Date(), "sprat_inverts_cleaned.csv", sep = "_")))
```





